Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest

### 2020-07-17 manuscript/climate_sensitivity_draft.Rmd
updating discussion
#112

Tree height, microenvironment, and hydraulic traits shaped tree growth responses across three droughts at our study site (Table 1, Fig. **4**). The greater susceptibility of larger trees to drought, similar to forests worldwide [@bennett_larger_2015], was driven primarily by their height [@stovall_tree_2019]. *Height was asscociated with...* The negative effect of height on $Rt$ held after accounting for species' traits. There was no evidence that soil water availability increased drought resistance; in contrast, trees in wetter topographic positions had lower $Rt$ [@zuleta_drought-induced_2017; @stovall_tree_2019], and the larger potential rooting volume of large trees provided no advantage in the drier microenvironments. Drought resistance was not linked to species' $LMA$, wood density, or xylem type (ring- vs. diffuse porous), but was negatively correlated with leaf hydraulic traits ($PLA_{dry}$, $\pi_{tlp}$). This is the first report to our knowledge linking $PLA_{dry}$ and $\pi_{tlp}$ to growth reduction during drought. The directions of these responses were consistent across droughts, supporting the premise that they were driven by fundamental physiological mechanisms. However, the strengths of each predictor varied across droughts (Fig. **4**; Tables S6-S7), indicating that drought characteristics interact with tree size, microenvironment, and traits to shape which individuals are most affected. These findings advance our knowledge of the factors that make trees vulnerable to growth declines during drought--and, by extension, likely make them more vulnerable to mortality [@sapes_plant_2019].

@@

Our analysis indicates that tree height has an overall stronger influence on drought response than does crown position (Table 1). This is consistent with, and reinforces, previous findings that biophysical constraints make it impossible for trees to efficiently transport water to great heights and simultaneously maintain strong resistance and resilience to drought-induced embolism [@olson_plant_2018; @couvreur_water_2018; @roskilly_conflicting_2019]. However, the collinearity between the two variables (Fig. **3d**) makes it impossible to confidently partition causality. Taller trees are more likely to be in dominant canopy positions (Fig. **3d**) and, largely as a consequence of their position relative to others, face different microenvironments (Fig. **3a-b**). Even under non-drought conditions, evaporative demand and maximum leaf temperatures increase with tree height [@smith_temperature_1977; @bretfeld_plant_2018; @kunert_revised_2017], and such conditions would incur additional stress during drought, when solar radiation tends to be higher and less water is available for evaporative cooling of the leaves (*Campbell & Norman REF*). *Previous studies have shown lower drought resistance in more exposed trees [@liu_effect_1993; @suarez_factors_2004; @scharnweber_confessions_2019], and ...* However, some decoupling between height and crown position is introduced by the configuration of neighboring trees (Fig. **3d**) [@muller-landau_testing_2006], and height was an overall stronger predictor of drought response than crown position (Fig. **4**; Tables 1). Belowground, taller trees would tend to have larger root systems, but the potentially greater access to water did not override the disadvantage conferred by height--and, in fact, greater moisture access in non-drought years (here, higher TWI) appears to make trees more sensitive to drought [@zuleta_drought-induced_2017; @stovall_tree_2019].

*Our analysis has the limitation that crown positions were recorded in 2018, as opposed to the years of the droughts. However, because trees would generally advance towards more dominant positions as they grow and as neighbors die, changing crown positions would bias against the acceptance of our hypothesis. The implication is that dominant crown positions did have a marginally negative influence on $Rt$, which makes sense in light of the vertical environmental gradients described above and agrees with previous studies showing lower drought resistance in more exposed trees [@liu_effect_1993; @suarez_factors_2004; @scharnweber_confessions_2019]. It is safe to assume that currently suppressed trees were suppressed throughout our analysis period, and their relatively low $Rt$ (after accounting for height effects) is real, perhaps as a result of competition [@sohn_potential_2016]. The observed height-sensitivity of $Rt$, together with the lack of advantage to large stature in drier topographic positions, agrees with the concept that physiological limitations to transpiration under drought shift from soil water availability to the plant-atmosphere interface as forests age [@bretfeld_plant_2018], such that tall, dominant trees are the most sensitive in mature forests. Additional research comparing drought responses of young and old forest stands, along with short and tall isolated trees, would be valuable for more clearly disentangling the roles of tree height and crown exposure. *

>>>>

Tree height, microenvironment, and hydraulic traits shaped tree growth responses across three droughts at our study site (Table 1, Fig. **4**). The greater susceptibility of larger trees to drought, similar to forests worldwide [@bennett_larger_2015], was driven primarily by their height [@stovall_tree_2019]. Taller height was likely a liability in itself, and was also associated with greater exposure to conditions that would promote water loss and heat damage during drought (Fig. **3**). There was no evidence that soil water availability increased drought resistance; in contrast, trees in wetter topographic positions had lower $Rt$ [@zuleta_drought-induced_2017; @stovall_tree_2019], and the larger potential rooting volume of large trees provided no advantage in the drier microenvironments. The negative effect of height on $Rt$ held after accounting for species' traits. Drought resistance was not consistently linked to species' $LMA$, wood density, or xylem type (ring- vs. diffuse porous), but was negatively correlated with leaf hydraulic traits ($PLA_{dry}$, $\pi_{tlp}$). This is the first report to our knowledge linking $PLA_{dry}$ and $\pi_{tlp}$ to growth reduction during drought. The directions of these responses were consistent across droughts (Table S6), supporting the premise that they were driven by fundamental physiological mechanisms. However, the strengths of each predictor varied across droughts (Fig. **4**; Tables S6-S7), indicating that drought characteristics interact with tree size, microenvironment, and traits to shape which individuals are most affected. These findings advance our knowledge of the factors that make trees vulnerable to growth declines during drought--and, by extension, likely make them more vulnerable to mortality [@sapes_plant_2019].

@@

Consistent with studies in other forests worldwide (**REFS**), taller trees in this forest exhibited lower drought resistance. 
Mechanistically, this is consistent with, and reinforces, previous findings that biophysical constraints make it impossible for trees to efficiently transport water to great heights and simultaneously maintain strong resistance and resilience to drought-induced embolism [@olson_plant_2018; @couvreur_water_2018; @roskilly_conflicting_2019]. 
Taller trees also face dramatically different microenvironments (Fig. **3**). 
They are exposed to higher wind speeds and lower humidity (Fig. **3a-b**), resulting in higher evaporative demand (**REFS**). 
Unlike other temperate forests where modestly cooler understory conditions have been documented (**REFS**), particularly under drier conditions (**REFS**), we observed no significant variation in air temperatures across the vertical profile (Fig. **3c**). 
More critically for tree physiology, leaf temperatures can become significantly elevated over air temperature under conditions of high solar radiation and low stomatal conductance (*Campbell & Norman REF*). 
Under drought, when air temperatures tend to be warmer, direct solar radiation tends to be higher (because of less cloud cover), and less water is available for evaporative cooling of the leaves, trees with sun-exposed crowns may not be able to simultaneously maintain leaf temperatures below damaging extremes and avoid drought-induced embolism. 
Indeed previous studies have shown lower drought resistance in more exposed trees [@liu_effect_1993; @suarez_factors_2004; @scharnweber_confessions_2019].
Unfortunately, collinearity between height and crown exposure in this study (Fig. **3d**) makes it impossible to confidently partition causality. 
Additional research comparing drought responses of early successional and mature forest stands, along with short and tall isolated trees, would be valuable for more clearly disentangling the roles of tree height and crown exposure.   
Belowground, taller trees would tend to have larger root systems, but the potentially greater access to water did not override the disadvantage conferred by height--and, in fact, greater moisture access in non-drought years (here, higher TWI) appears to make trees more sensitive to drought [@zuleta_drought-induced_2017; @stovall_tree_2019].
The observed height-sensitivity of $Rt$, together with the lack of advantage to large stature in drier topographic positions, agrees with the concept that physiological limitations to transpiration under drought shift from soil water availability to the plant-atmosphere interface as forests age [@bretfeld_plant_2018], such that tall, dominant trees are the most sensitive in mature forests. 
Again, additional research comparing drought responses across forests with different tree heights and water availability would be valuable for disentangling the relative importance of above- and belowground mechanisms.

*Even under non-drought conditions, evaporative demand and maximum leaf temperatures increase with tree height [@smith_temperature_1977; @bretfeld_plant_2018; @kunert_revised_2017], and such conditions would incur additional stress during drought, when solar radiation tends to be higher and less water is available for evaporative cooling of the leaves .*
